These last five years would not have been possible without the help of many. To all who have helped in any way, thank you.

First off, I thank my supervisor, Thomas Lumley, for proposing the entire context of this thesis, and for five years being available (most of the time; there were one or two overseas conferences) for my usually unannounced drop-ins.
%Thomas has a knack of providing very clear solutions to problems, often making complicated concepts seem obvious (there were many internal exclamations of ``oh duh!'').
I also thank Brendon Brewer, my co-supervisor, for early help getting started with \Cpp{}, and for offering his assistance should ever I need it.



To the entire statistics department at the University of Auckland: thank you for being so welcoming and supportive for the past five years (and more), in particular to Chris Wild for your incredible support of both my current work and future career opportunities. Thank you to the university itself for providing me with the opportunity to pursue doctoral study, and for the 3.5 years of financial support through the Doctoral Scholarship.
And to the Center for e-Research for providing a \gls{vm} for my use, which allowed  me to test the real-time feasibility of what I was trying to do. Thank you to the individuals involved in providing me with these much-needed resources.

To my friends both within and without the department, thank you for providing humour, social escapes, and those essential reminders that ``we all feel the same way, it's normal!''.

To my delightful partner Mark, who has been supporting a broke student while he frantically condenses 4~years' work into 200~pages. For your emotional support over recent months, and the occasional nudge to get things done: thank you.

And lastly, to the two most loving and supportive parents one could want. These last two years have been tough, but despite the challenges you were there for me whenever I needed a ride, a roof over my head, or a home-cooked meal. My success is a testament to your unremitting love and support. Thank you.
