\subsection{Estimating dwell times at stops}
\label{sec:trip_state_dwell}


The dwell time at stops is also an unknown parameter
that accounts for a lot of the variability.
At peak times, buses can spend a lot of time picking up
and dropping off passengers at all stops between
where it is now, and the target bus stop,
all of which can add up.
However, it is also possible that,
for one reason or another,
no passengers embark or disembark the bus,
and it needn't stop at all.
We need to be able to account for this in the arrival time estimates.

NOTE TO SELF: Eq (3.11) on page 15 needs to be fixed - p (gamma + tau) is wrong ... truncation point should be 0

We described bus stop dwell time in \cref{sec:vehicle_model_stops},
and the distribution is shown in
\cref{fig:eta_dwell_times} on \cpageref{fig:eta_dwell_times}.
The model for dwell time is
\begin{gather}
\label{eq:stop_dwell_times}
\Tdwell_{j} = \Istop_{j} (\mindwell + \dot D_{j}) \\
\dot D_{j} \sim \TNormal{\dwell_{j}}{\dwellvar_{j}}{0}{\infty} \nonumber \\
\Istop_{j} \sim \Bern{\Prstop_{j}} \nonumber
\end{gather}

So we can compute the expectation and variance of this thing.
First, we assume that dwell time and stopping probability are independent,
which is likely not far from the truth.
There is, of course, relationship between the two,
but it is very complicated and depends more so on other buses
that the parameters themselves.
So, we can trivially show that the expected dwell time is
\begin{align}
\label{eq:trip_dwell_mean}
\E{\Tdwell_{j}}
&= \E{\Istop_{j} (\mindwell + \dot D_{j})} \nonumber \\
&= \E{\Istop_{j} \mindwell} + \E{\Istop_{j}\dot D_{j}} \nonumber \\
&= \Prstop_{j}\mindwell + \Prstop_{j}\dwell_{j}
\end{align}
with variance
\begin{align}
\label{eq:trip_dwell_var}
\Var{\Tdwell_{j}}
&= \Var{\Istop_{j} (\mindwell + \dot D_{j})} \nonumber \\
&= \Big(\Var{\Istop_{j}} + \E{\Istop_{j}}^2\Big)
    \Big(\Var{\mindwell + \dot D_{j}} + \E{\mindwell + \dot D_{j}}^2\Big)
    \nonumber \\
    &\phantom{=} - \E{\Istop\vphantom{\dot D_{j}}}^2 \E{\mindwell + \dot D_{j}}^2 \nonumber \\
&= \left(\Prstop_{j}(1-\Prstop_{j}) + \Prstop_{j}^2\right)
    \left(\dwellvar_{j} + (\mindwell + \dwell_{k})^2\right) -
    \Prstop_{j}^2 (\mindwell + \dwell_{j})^2 \nonumber  \\
&= \Prstop_{j}(\dwellvar + (\mindwell + \dwell_{j})^2) -
    \Prstop_{j}^2(\mindwell + \dwell_{j})^2  \nonumber \\
&= \Prstop_{j} \left(
    \dwellvar_{j} + (1-\Prstop_{j})(\mindwell + \dwell_{j})^2
\right)
\end{align}



While this is certainly not a good approximation of dwell time distributions,
it makes it easy to combine to travel times and other stop dwell times
to obtain a final arrival time distribution.

